---
title: "access_lowell"
author: "SP"
date: "2/12/2022"
output: html_document
---

```{r message = FALSE, warning=FALSE, echo=FALSE}
options(java.parameters = "-Xmx3G")

library(r5r)
library(osmextract)
library(tidyverse)
library(sf)
library(ggthemes)
library(ggspatial)
library(tigris)
library(wesanderson)
library(tidytransit)
library(stars)

#dir.create("networks")
```

```{r}
rta_stops <- st_read("RTA_Bus_Stops")

test_plot <- ggplot(rta_stops) +
  geom_sf()

test_plot
```

```{r message = FALSE, warning=FALSE, echo=FALSE}
Lowell_file <- oe_match("Lowell, Massachusetts")
```

```{r message = FALSE, warning=FALSE, echo=FALSE}
Lowell_streets <- oe_read(Lowell_file$url, 
                   provider = "geofabrik", 
                   download_directory = "networks", 
                   layer = "lines", 
                   quiet = TRUE) %>%
  filter(!is.na(highway)) 
```

```{r message = FALSE, warning=FALSE, echo=FALSE}
MA_state_plane <- "+proj=lcc +lat_1=41.71666666666667 +lat_2=42.68333333333333 +lat_0=41 +lon_0=-71.5 +x_0=200000 +y_0=750000 +ellps=GRS80 +units=m +no_defs"

Lowell_city_limits <- places("Massachusetts") %>%
  filter(NAME == "Lowell") 

Lowell_city_limits <- Lowell_city_limits %>%
  st_transform(MA_state_plane)

Lowell_streets <- Lowell_streets %>%
  st_transform(MA_state_plane)

rta_stops <- rta_stops %>%
  st_transform(MA_state_plane)
```

```{r message = FALSE, warning=FALSE, echo=FALSE}
Lowell_streets <- Lowell_streets[Lowell_city_limits,]

Lowell_stops <- rta_stops[Lowell_city_limits,]

ggplot(Lowell_streets) +
  geom_sf(color = 'gray') +
  geom_sf(data = Lowell_stops, color = "darkblue") +
  coord_sf(crs = MA_state_plane)  +
  theme_void()
```

```{r message = FALSE, warning=FALSE, echo=FALSE}
grid <- st_sf(st_make_grid(Lowell_city_limits, 
                           square = FALSE, 
                           n = c(100,100),
                           what = "polygons")) %>%
  st_filter(Lowell_city_limits) 

colnames(grid) <- "geometry"
st_geometry(grid) <- "geometry"

grid <- grid %>%
  mutate(id = seq(1, length(grid$geometry), by=1))

Lowell_stops <- Lowell_stops %>% 
mutate(id = seq(1, length(Lowell_stops$geometry), by = 1))

grid_points <- st_centroid(grid)
```

```{r}
ggplot(grid) +
  geom_sf(color = 'gray') +
  geom_sf(data = Lowell_stops, color = "darkblue") +
  geom_sf(data = Lowell_streets, alpha = 0.2)
  coord_sf(crs = MA_state_plane)
```


```{r message = FALSE, warning=FALSE, echo=FALSE}
r5r_core <- setup_r5("networks", verbose = FALSE)
```

```{r message = FALSE, warning=FALSE, echo=FALSE}
ttm <- travel_time_matrix(r5r_core = r5r_core,
                          origins = Lowell_stops,
                          destinations = grid_points,
                          mode = "WALK",
                          max_trip_duration = 40,
                          verbose = FALSE)
```

```{r message = FALSE, warning=FALSE, echo=FALSE}
ttm__wide <- ttm %>%
  pivot_wider(names_from = fromId, 
              names_prefix = "from", values_from = travel_time) %>%
  rename(id = toId) %>% 
  merge(grid) %>%
  replace(is.na(.), 999) %>%
  rowwise() %>%
  mutate(from_any = min(c_across(starts_with("from")), na.rm = TRUE))

st_geometry(ttm_wide) <- "geometry"
```


 
```{r message = FALSE, warning=FALSE, echo=FALSE}
iso_palette <- wes_palette("FantasticFox1", n = 5)

iso10min <- ttms_wide[ttm_wide$from_any < 11,] %>%
  st_union()

iso20min <- ttm_wide[ttm_wide$from_any < 21,] %>%
  st_union()

iso30min <- ttm_wide[ttm_wide$from_any < 31,] %>%
  st_union()

ggplot(Lowell_streets) +
  geom_sf(data = iso30min, 
          aes(fill = "Area within 30 minutes"), 
          color = NA) +
  geom_sf(data = iso20min, 
          aes(fill = "Area within 20 minutes"), 
          color = NA) +
  geom_sf(data = iso10min, 
          aes(fill = "Area within 10 minutes"), 
          color = NA) +
  geom_sf(alpha = 0.5) +
  scale_fill_manual(values = c(iso_palette[3], 
                               iso_palette[1],
                               iso_palette[5]),
        name = "Walk travel\ntime to the\nnearest bus stop\n(minutes)") +
  coord_sf(crs = MA_state_plane) +
  theme_map() + 
  theme(legend.position = c(0.75, 0.70))
```


Map 2 illustrates the cutoff travel times via walking to bus stops in Lowell, separated into bins of 10, 20 and 30 minutes. We’re pleased to find that most of Lowell is serviced by a bus stop within a 10-minute walking window. 


```{r message = FALSE, warning=FALSE, echo=FALSE}
iso10min_b <- tt_pschools_wide[tt_bustop_wide$from_any < 11,] %>%
  st_union()

iso20min_b <- tt_pschools_wide[tt_bustop_wide$from_any < 21,] %>%
  st_union()

iso30min_b <- tt_pschools_wide[tt_bustop_wide$from_any < 31,] %>%
  st_union()

ggplot(Lowell_streets) +
  geom_sf(data = iso30min_b, 
          aes(fill = "Area within 30 minutes"), 
          color = NA) +
  geom_sf(data = iso20min_b, 
          aes(fill = "Area within 20 minutes"), 
          color = NA) +
  geom_sf(data = iso10min_b, 
          aes(fill = "Area within 10 minutes"), 
          color = NA) +
  geom_sf(alpha = 0.5) +
  scale_fill_manual(values = c(iso_palette[3], 
                               iso_palette[1],
                               iso_palette[5]),
        name = "Walk travel\ntime to the\nnearest bus stop\n(minutes)") +
  coord_sf(crs = MA_state_plane) +
  theme_map()
```





**2. A set of points, polygons, or contours throughout the city with colors varying continuously to indicate the travel time from each point to the nearest location of interest.**

 
Map 3 illustrates walking times to public schools in Lowell on a continuous color scale.  A continuous scale shows more accurate travel times because the scale is more than the 3 bins used in Map 1. Additionally, it provides a more thorough picture of the city of Lowell as it covers areas that have a travel time of more than 30 minutes.  


```{r message = FALSE, warning=FALSE, echo=FALSE}
ttw_palette <- wes_palette("Zissou1", n = 5)

ggplot(Lowell_streets) +
  geom_sf(data = tt_pschools_wide, 
          aes(fill = from_any), 
          color = NA) +
  geom_sf(alpha = 0.5) +
  scale_fill_gradient2(low = ttw_palette[1], mid = ttw_palette[3], high = ttw_palette[5], 
                       midpoint = 30,
        name = "Walk travel\ntime to the\nnearest public school\n(minutes)") +
  coord_sf(crs = MA_state_plane) +
  theme_map()
```



Map 4 illustrates walking times to bus stops in Lowell on a continuous color scale. A continuous scale shows more accurate travel times because the scale is more than the 3 bins used in Map 1. However, because there are generally many bus stops across Lowell within 10 minutes of each other, the maps are overall similar. 



```{r message = FALSE, warning=FALSE, echo=FALSE}
ggplot(Lowell_streets) +
  geom_sf(data = tt_bustop_wide, 
          aes(fill = from_any), 
          color = NA) +
  geom_sf(alpha = 0.5) +
  scale_fill_gradient2(low = ttw_palette[1], mid = ttw_palette[3], high = ttw_palette[5], 
                       midpoint = 30,
        name = "Walk travel\ntime to the\nnearest bus stop\n(minutes)") +
  coord_sf(crs = MA_state_plane) +
  theme_map()
```
```{r message = FALSE, warning=FALSE, echo=FALSE}
pschool_grid <- grid %>%
  mutate(num_pschools = lengths(st_covers(grid, Lowell_pschools)))

pschool_points <- st_centroid(pschool_grid)
```

```{r message = FALSE, warning=FALSE, echo=FALSE}
busstop_grid <- grid %>%
  mutate(num_busstops = lengths(st_covers(grid, Lowell_busstop)))

busstop_points <- st_centroid(busstop_grid)
```

```{r message = FALSE, warning=FALSE, echo=FALSE}
Lowell_access_pschools <- accessibility(r5r_core,
                        origins = pschool_points,
                        destinations = pschool_points,
                        mode = "WALK",
                        opportunities_colname = "num_pschools",
                        decay_function = "exponential",
                        cutoffs = 5,
                        time_window = 60,
                        percentiles = 50,
                        verbose = FALSE) %>%
  mutate(id = as.numeric(from_id)) %>%
  merge(grid)
```     

```{r message = FALSE, warning=FALSE, echo=FALSE}
Lowell_access_busstop <- accessibility(r5r_core,
                        origins = busstop_points,
                        destinations = busstop_points,
                        mode = "WALK",
                        opportunities_colname = "num_busstops",
                        decay_function = "exponential",
                        cutoffs = 5,
                        time_window = 60,
                        percentiles = 50,
                        verbose = FALSE) %>%
  mutate(id = as.numeric(from_id)) %>%
  merge(grid)
```

```{r message = FALSE, warning=FALSE, echo=FALSE}
stop_r5(r5r_core)

rJava::.jgc(R.gc = TRUE)

st_write(Lowell_access_pschools, 'Lowell_access_pschools.geojson', append=FALSE, quiet=TRUE )

st_write(Lowell_access_busstop, 'Lowell_access_busstop.geojson', append=FALSE, quiet=TRUE)
```



**4. A set of points, polygons, or contours throughout the city with values varying continuously to indicate an accessibility score to locations of interest, based on a continuous decay function.**


Map 5 demonstrates accessibility using raster data and displays the accessibility of public schools in Lowell using a score assigned by the r5r package. Raster data is helpful for this map because you can compare two adjacent points and their differing accessibilities. 

 
```{r message = FALSE, warning=FALSE, echo=FALSE}
access_poly_p <- st_read("Lowell_access_pschools.geojson", quiet=TRUE)

access_raster_p <- st_rasterize(access_poly_p["accessibility"], 
                              nx = 100, ny = 100) 
plot(access_raster_p)
```



Map 6 demonstrates accessibility using raster data and displays the accessibility of bus stops in Lowell using a score assigned by the r5r package. Raster data is helpful for this map because you can compare two adjacent points and their differing accessibilities. 



```{r message = FALSE, warning=FALSE, echo=FALSE}
access_poly_b <- st_read("Lowell_access_busstop.geojson", quiet=TRUE)

access_raster_b <- st_rasterize(access_poly_b["accessibility"], 
                              nx = 100, ny = 100) 
plot(access_raster_b)
```



Map 7 demonstrates pedestrian access to schools by converting the raster layer to a vector layer and plotting the data as points.  


 
```{r message = FALSE, warning=FALSE, echo=FALSE}
access_points_p <- st_as_sf(access_raster_p, as_points = TRUE)

ggplot(Lowell_streets) +
  geom_sf(data = access_points_p, aes(color = accessibility), size = 1) +
  scale_color_viridis_c(na.value = NA, 
                       option="D",
                       name = "Pedestrian access to\npublic schools") +
  theme_void() + 
  theme(legend.position = c(0.75, 0.70))
```


Map 8 demonstrates pedestrian access to bus stops by converting the raster layer to a vector layer and plotting the data as points. This map displays a high concentration of access in the middle of Lowell. 


```{r message = FALSE, warning=FALSE, echo=FALSE}
access_points_b <- st_as_sf(access_raster_b, as_points = TRUE)

ggplot(Lowell_streets) +
  geom_sf(data = access_points_b, aes(color = accessibility), size = 1) +
  scale_color_viridis_a(na.value = NA, 
                       option="A",
                       name = "Pedestrian access to\nbus stops") +
  theme_void()
```


**Contribution Statement**

We allocate 10 points to each group member for their work on this assignment. Samantha found the public school points and Naomi found the bus stop data. The group collectively troubleshooted the code using the various tutorials available. Manoel and Naomi wrote the descriptions. Samantha knit the final file. 
